quollr: An R Package for Visualizing 2-D Models from Non-linear Dimension Reductions in High Dimensional Space

### 2025-05-15 paper-quollr.Rmd
Updated link plot texts

Before visualization, the input data must be prepared using the `comb_all_data_model()` function. This function combines the high-dimensional data (`highd_data`), the NLDR output (`nldr_data`), and the bin-averaged high-dimensional model representation (`model_highd`) aligned to the 2D bin layout (`model_2d`):

@@

The function `show_link_plots()` generates two side-by-side, interactively linked plots; a 2D scatterplot of the NLDR representation, and a dynamic projection tour in the original high-dimensional space (using the `langevitour` package), displaying both the data and the model.

@@

`show_error_link_plots()` helps to see investigate whether the model fits the points everywhere or fits better in some places, or simply mismatches the pattern. The function requires several arguments: points data which contain Non-linear dimension reduction data, high-dimensional data, high-dimensional model data, and model error, and edge data where the from and to links of the edges.

```{r, echo=TRUE}
df_exe <- comb_all_data_model_error(
  highd_data = scurve, 
  nldr_data = scurve_umap, 
  model_highd = df_bin, 
  model_2d = df_bin_centroids, 
  error_data = model_error
  )
```

@@

# Find the most appropriate fit

```{r}
#| label: scurve-diff-umap-layouts

nldr_scaled_obj <- gen_scaled_data(
  nldr_data = scurve_umap)
umap_scaled1 <- nldr_scaled_obj$scaled_nldr

scurve_nldr1 <- umap_scaled1 |>
  ggplot(aes(x = emb1,
             y = emb2))+
  geom_point(alpha=0.1, size=1, colour='#e41a1c') +
  interior_annotation("a", c(0.08, 0.9))

nldr_scaled_obj2 <- gen_scaled_data(
  nldr_data = scurve_umap2)
umap_scaled2 <- nldr_scaled_obj2$scaled_nldr

scurve_nldr2 <- umap_scaled2 |>
  ggplot(aes(x = emb1,
             y = emb2))+
  geom_point(alpha=0.1, size=1, colour='#ff7f00') +
  interior_annotation("b")

nldr_scaled_obj3 <- gen_scaled_data(
  nldr_data = scurve_umap3)
umap_scaled3 <- nldr_scaled_obj3$scaled_nldr

scurve_nldr3 <- umap_scaled3 |>
  ggplot(aes(x = emb1,
             y = emb2))+
  geom_point(alpha=0.1, size=1, colour='#4daf4a') +
  interior_annotation("c", c(0.08, 0.9))

nldr_scaled_obj4 <- gen_scaled_data(
  nldr_data = scurve_umap4)
umap_scaled4 <- nldr_scaled_obj4$scaled_nldr

scurve_nldr4 <- umap_scaled4 |>
  ggplot(aes(x = emb1,
             y = emb2))+
  geom_point(alpha=0.1, size=1, colour='#a65628') +
  interior_annotation("d")

nldr_scaled_obj5 <- gen_scaled_data(
  nldr_data = scurve_umap5)
umap_scaled5 <- nldr_scaled_obj5$scaled_nldr

scurve_nldr5 <- umap_scaled5 |>
  ggplot(aes(x = emb1,
             y = emb2))+
  geom_point(alpha=0.1, size=1, colour='#636363') +
  interior_annotation("e")

nldr_scaled_obj6 <- gen_scaled_data(
  nldr_data = scurve_umap6)
umap_scaled6 <- nldr_scaled_obj6$scaled_nldr

scurve_nldr6 <- umap_scaled6 |>
  ggplot(aes(x = emb1,
             y = emb2))+
  geom_point(alpha=0.1, size=1, colour='#984ea3') +
  interior_annotation("f")
```

```{r}
scurve_error <- dplyr::bind_rows(scurve_umap_mse,
                                 scurve_umap_mse2,
                                 scurve_umap_mse3,
                                 scurve_umap_mse4,
                                 scurve_umap_mse5,
                                 scurve_umap_mse6)

scurve_error <- scurve_error |>
  mutate(a1 = round(a1, 2)) |>
  filter(bin1 >= 5) |>
  group_by(method, a1) |>
  filter(MSE == min(MSE)) |>
  ungroup()

error_plot <- ggplot(scurve_error,
         aes(x = a1,
             y = sqrt(MSE),
             color = method)) +
    geom_point(size = 0.8) +
    geom_line(linewidth = 0.3) +
  scale_x_continuous(breaks =
    sort(unique(scurve_error$a1))[
      seq(1, length(
        unique(scurve_error$a1)), 
        by = 5)]) +
  scale_color_manual(
    values=c('#e41a1c','#ff7f00','#4daf4a', 
             "#a65628",'#636363', '#984ea3'))+
    ylab("RMSE") +
    xlab(expression(paste("binwidth (", a[1], ")"))) +
    theme_minimal() +
    theme(panel.border = element_rect(fill = 'transparent'),
          plot.title = element_text(size = 12, hjust = 0.5, vjust = -0.5),
          axis.ticks.x = element_line(),
          axis.ticks.y = element_line(),
          legend.position = "none",
          axis.text.x = element_text(size = 7),
          axis.text.y = element_text(size = 7),
          axis.title.x = element_text(size = 7),
          axis.title.y = element_text(size = 7),
          aspect.ratio = 1.5)
```

```{r}
free(error_plot) + wrap_plots(
  scurve_nldr1, scurve_nldr2, 
  scurve_nldr3, scurve_nldr4,
  scurve_nldr5, scurve_nldr6, 
  ncol = 2)
```

>>>>

Before visualization, the input data must be prepared using the `comb_all_data_model()` function. This function combines the high-dimensional data (`highd_data`), the NLDR data (`nldr_data`), and the bin-averaged high-dimensional model representation (`model_highd`) aligned to the $2-\text{D}$ bin layout (`model_2d`):

@@

The function `show_link_plots()` generates two side-by-side, interactively linked plots; a $2-\text{D}$ NLDR representation, and a dynamic projection tour in the original high-dimensional space (using the `langevitour` package), displaying both the data and the model. The function takes the output from `comb_all_data_model()` (`point_data`) and `edge_data` which defines connections between neighboring bins.

These two views are linked using `crosstalk`, allowing interactive selection of points in the NLDR plot to highlight corresponding structures in the `langevitour` output.

`show_error_link_plots()` helps to see investigate whether the model fits the points everywhere or fits better in some places, or simply mismatches the pattern.

Before visualization, the input data must be prepared using the `comb_all_data_model_error()` function. The function requires several arguments: points data which contain high-dimensional data (`highd_data`), NLDR data (`nldr_data`), high-dimensional model data (`model_highd`), $2-\text{D}$ model data (`model_2d`), and model error (`error_data`).

```{r, echo=TRUE}
df_exe <- comb_all_data_model_error(
  highd_data = scurve, 
  nldr_data = scurve_umap, 
  model_highd = df_bin, 
  model_2d = df_bin_centroids, 
  error_data = model_error
  )

The function `show_error_link_plots()` generates three side-by-side, interactively linked plots; a error distribution, a $2-\text{D}$ NLDR representation, and a dynamic projection tour in the original high-dimensional space (using the `langevitour` package), displaying both the data and the model. The function takes the output from `comb_all_data_model_error()` (`point_data`) and `edge_data` which defines connections between neighboring bins.

@@

These three views are linked using `crosstalk`, allowing interactive selection of points in error plot and the NLDR plot to highlight corresponding structures in the `langevitour` output.

@@

<!-- # Find the most appropriate fit -->

<!-- ```{r} -->
<!-- #| label: scurve-diff-umap-layouts -->

<!-- nldr_scaled_obj <- gen_scaled_data( -->
<!--   nldr_data = scurve_umap) -->
<!-- umap_scaled1 <- nldr_scaled_obj$scaled_nldr -->

<!-- scurve_nldr1 <- umap_scaled1 |> -->
<!--   ggplot(aes(x = emb1, -->
<!--              y = emb2))+ -->
<!--   geom_point(alpha=0.1, size=1, colour='#e41a1c') + -->
<!--   interior_annotation("a", c(0.08, 0.9)) -->

<!-- nldr_scaled_obj2 <- gen_scaled_data( -->
<!--   nldr_data = scurve_umap2) -->
<!-- umap_scaled2 <- nldr_scaled_obj2$scaled_nldr -->

<!-- scurve_nldr2 <- umap_scaled2 |> -->
<!--   ggplot(aes(x = emb1, -->
<!--              y = emb2))+ -->
<!--   geom_point(alpha=0.1, size=1, colour='#ff7f00') + -->
<!--   interior_annotation("b") -->

<!-- nldr_scaled_obj3 <- gen_scaled_data( -->
<!--   nldr_data = scurve_umap3) -->
<!-- umap_scaled3 <- nldr_scaled_obj3$scaled_nldr -->

<!-- scurve_nldr3 <- umap_scaled3 |> -->
<!--   ggplot(aes(x = emb1, -->
<!--              y = emb2))+ -->
<!--   geom_point(alpha=0.1, size=1, colour='#4daf4a') + -->
<!--   interior_annotation("c", c(0.08, 0.9)) -->

<!-- nldr_scaled_obj4 <- gen_scaled_data( -->
<!--   nldr_data = scurve_umap4) -->
<!-- umap_scaled4 <- nldr_scaled_obj4$scaled_nldr -->

<!-- scurve_nldr4 <- umap_scaled4 |> -->
<!--   ggplot(aes(x = emb1, -->
<!--              y = emb2))+ -->
<!--   geom_point(alpha=0.1, size=1, colour='#a65628') + -->
<!--   interior_annotation("d") -->

<!-- nldr_scaled_obj5 <- gen_scaled_data( -->
<!--   nldr_data = scurve_umap5) -->
<!-- umap_scaled5 <- nldr_scaled_obj5$scaled_nldr -->

<!-- scurve_nldr5 <- umap_scaled5 |> -->
<!--   ggplot(aes(x = emb1, -->
<!--              y = emb2))+ -->
<!--   geom_point(alpha=0.1, size=1, colour='#636363') + -->
<!--   interior_annotation("e") -->

<!-- nldr_scaled_obj6 <- gen_scaled_data( -->
<!--   nldr_data = scurve_umap6) -->
<!-- umap_scaled6 <- nldr_scaled_obj6$scaled_nldr -->

<!-- scurve_nldr6 <- umap_scaled6 |> -->
<!--   ggplot(aes(x = emb1, -->
<!--              y = emb2))+ -->
<!--   geom_point(alpha=0.1, size=1, colour='#984ea3') + -->
<!--   interior_annotation("f") -->
<!-- ``` -->

<!-- ```{r} -->
<!-- scurve_error <- dplyr::bind_rows(scurve_umap_mse, -->
<!--                                  scurve_umap_mse2, -->
<!--                                  scurve_umap_mse3, -->
<!--                                  scurve_umap_mse4, -->
<!--                                  scurve_umap_mse5, -->
<!--                                  scurve_umap_mse6) -->

<!-- scurve_error <- scurve_error |> -->
<!--   mutate(a1 = round(a1, 2)) |> -->
<!--   filter(bin1 >= 5) |> -->
<!--   group_by(method, a1) |> -->
<!--   filter(MSE == min(MSE)) |> -->
<!--   ungroup() -->

<!-- error_plot <- ggplot(scurve_error, -->
<!--          aes(x = a1, -->
<!--              y = sqrt(MSE), -->
<!--              color = method)) + -->
<!--     geom_point(size = 0.8) + -->
<!--     geom_line(linewidth = 0.3) + -->
<!--   scale_x_continuous(breaks = -->
<!--     sort(unique(scurve_error$a1))[ -->
<!--       seq(1, length( -->
<!--         unique(scurve_error$a1)),  -->
<!--         by = 5)]) + -->
<!--   scale_color_manual( -->
<!--     values=c('#e41a1c','#ff7f00','#4daf4a',  -->
<!--              "#a65628",'#636363', '#984ea3'))+ -->
<!--     ylab("RMSE") + -->
<!--     xlab(expression(paste("binwidth (", a[1], ")"))) + -->
<!--     theme_minimal() + -->
<!--     theme(panel.border = element_rect(fill = 'transparent'), -->
<!--           plot.title = element_text(size = 12, hjust = 0.5, vjust = -0.5), -->
<!--           axis.ticks.x = element_line(), -->
<!--           axis.ticks.y = element_line(), -->
<!--           legend.position = "none", -->
<!--           axis.text.x = element_text(size = 7), -->
<!--           axis.text.y = element_text(size = 7), -->
<!--           axis.title.x = element_text(size = 7), -->
<!--           axis.title.y = element_text(size = 7), -->
<!--           aspect.ratio = 1.5) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- free(error_plot) + wrap_plots( -->
<!--   scurve_nldr1, scurve_nldr2,  -->
<!--   scurve_nldr3, scurve_nldr4, -->
<!--   scurve_nldr5, scurve_nldr6,  -->
<!--   ncol = 2) -->
<!-- ``` -->

### 2025-05-15 paper-quollr.tex
Updated link plot texts

\begin{verbatim}
#> # A tibble: 1 x 2
#>   Error    MSE
#>   <dbl>  <dbl>
#> 1 1481. 0.0325
\end{verbatim}

@@

Before visualization, the input data must be prepared using the \texttt{comb\_all\_data\_model()} function. This function combines the high-dimensional data (\texttt{highd\_data}), the NLDR output (\texttt{nldr\_data}), and the bin-averaged high-dimensional model representation (\texttt{model\_highd}) aligned to the 2D bin layout (\texttt{model\_2d}):

@@

The function \texttt{show\_link\_plots()} generates two side-by-side, interactively linked plots; a 2D scatterplot of the NLDR representation, and a dynamic projection tour in the original high-dimensional space (using the \texttt{langevitour} package), displaying both the data and the model.

@@

\texttt{show\_error\_link\_plots()} helps to see investigate whether the model fits the points everywhere or fits better in some places, or simply mismatches the pattern. The function requires several arguments: points data which contain Non-linear dimension reduction data, high-dimensional data, high-dimensional model data, and model error, and edge data where the from and to links of the edges.

\begin{verbatim}
df_exe <- comb_all_data_model_error(
  highd_data = scurve, 
  nldr_data = scurve_umap, 
  model_highd = df_bin, 
  model_2d = df_bin_centroids, 
  error_data = model_error
  )
\end{verbatim}

@@

\section{Find the most appropriate fit}\label{find-the-most-appropriate-fit}

\includegraphics[width=1\linewidth]{paper-quollr_files/figure-latex/unnamed-chunk-33-1}

>>>>

\begin{verbatim}
#> # A tibble: 1 x 2
#>   Error  RMSE
#>   <dbl> <dbl>
#> 1 1481. 0.180
\end{verbatim}

@@

Before visualization, the input data must be prepared using the \texttt{comb\_all\_data\_model()} function. This function combines the high-dimensional data (\texttt{highd\_data}), the NLDR data (\texttt{nldr\_data}), and the bin-averaged high-dimensional model representation (\texttt{model\_highd}) aligned to the \(2-\text{D}\) bin layout (\texttt{model\_2d}):

@@

The function \texttt{show\_link\_plots()} generates two side-by-side, interactively linked plots; a \(2-\text{D}\) NLDR representation, and a dynamic projection tour in the original high-dimensional space (using the \texttt{langevitour} package), displaying both the data and the model. The function takes the output from \texttt{comb\_all\_data\_model()} (\texttt{point\_data}) and \texttt{edge\_data} which defines connections between neighboring bins.

These two views are linked using \texttt{crosstalk}, allowing interactive selection of points in the NLDR plot to highlight corresponding structures in the \texttt{langevitour} output.

\texttt{show\_error\_link\_plots()} helps to see investigate whether the model fits the points everywhere or fits better in some places, or simply mismatches the pattern.

Before visualization, the input data must be prepared using the \texttt{comb\_all\_data\_model\_error()} function. The function requires several arguments: points data which contain high-dimensional data (\texttt{highd\_data}), NLDR data (\texttt{nldr\_data}), high-dimensional model data (\texttt{model\_highd}), \(2-\text{D}\) model data (\texttt{model\_2d}), and model error (\texttt{error\_data}).

\begin{verbatim}
df_exe <- comb_all_data_model_error(
  highd_data = scurve, 
  nldr_data = scurve_umap, 
  model_highd = df_bin, 
  model_2d = df_bin_centroids, 
  error_data = model_error
  )

\begin{verbatim}
#> # A tibble: 5,122 x 12
#>         x1    x2    x3         x4        x5       x6        x7 type   emb1  emb2
#>      <dbl> <dbl> <dbl>      <dbl>     <dbl>    <dbl>     <dbl> <chr> <dbl> <dbl>
#>  1 -0.992   1.91 1.11  -0.000427   0.000624  0.00749  0.00105  model    NA    NA
#>  2 -0.906   1.93 1.41  -0.0000183  0.00331  -0.0204  -0.000363 model    NA    NA
#>  3 -0.680   1.93 1.72  -0.000810  -0.00259  -0.00449  0.00153  model    NA    NA
#>  4  0.461   1.93 1.89  -0.00478    0.00492   0.00835  0.00172  model    NA    NA
#>  5 -0.985   1.75 0.853 -0.00202    0.000397  0.00331  0.000338 model    NA    NA
#>  6 -0.980   1.66 1.17  -0.000374  -0.00154   0.0165   0.000126 model    NA    NA
#>  7 -0.821   1.64 1.56  -0.000459   0.000538 -0.0123   0.000780 model    NA    NA
#>  8 -0.484   1.68 1.87   0.00313    0.00241   0.00823 -0.00117  model    NA    NA
#>  9 -0.0991  1.70 1.99   0.00103    0.00150   0.00877 -0.000193 model    NA    NA
#> 10  0.295   1.74 1.95  -0.00165    0.000459  0.00330  0.000257 model    NA    NA
#> # i 5,112 more rows
#> # i 2 more variables: sqrt_row_wise_total_error <dbl>, density <dbl>
\end{verbatim}

The function \texttt{show\_error\_link\_plots()} generates three side-by-side, interactively linked plots; a error distribution, a \(2-\text{D}\) NLDR representation, and a dynamic projection tour in the original high-dimensional space (using the \texttt{langevitour} package), displaying both the data and the model. The function takes the output from \texttt{comb\_all\_data\_model\_error()} (\texttt{point\_data}) and \texttt{edge\_data} which defines connections between neighboring bins.

@@

These three views are linked using \texttt{crosstalk}, allowing interactive selection of points in error plot and the NLDR plot to highlight corresponding structures in the \texttt{langevitour} output.

@@
